### 2015-03-12 design.tex
Minor edits to the text.
One note: as I read this, I realized that the reader doesn't need to
know that we have event and sensor PROTOTYPES instead of TYPES. That's
an implementation detail that comes from an oddity of PRAXIS. So I
suppressed mention of "prototypes."

% DONE: Can you give the reader a sense of the distinction between
% configuration, setting, and run? The notion of "run" is relatively clear, but
% not the other two.  Something like "For example, a configuration might specify
% how a set of attacks is generated, and the set of sensors. A particular
% setting would involve a particular set of possible attacks, a particular set
% of sensors, and a particular set of sensing relations between attacks and
% sensors."
%
% FIXME - Robert, in the marked-up PDF you suggested discussing
% locations, but earlier we'd decided to prune that discussion, since
% we didn't end up using different locations.
% Right, but here how will the reader know that we are going to sample from the
% same event multiple times?  The fact is that the whole notion of "run" as
% distinct from "location" is broken, so we must make the best of it that we
% can.  Please have a whack at this.
For each experiment we create a \emph{configuration},
parameters specifying how a set of events and sensors (the
\emph{setting}) is generated.  A
setting comprises a set of possible attacks, a set
of sensors, and a set of sensing relations between attacks and
sensors.
In each \emph{run} of a setting we sample %, for each location,
% some number of 
from the attack and benign events.
% These are
% Bernoulli random variables; we discuss setting their probabilities
% below.
In each experiment we generated 1,000 settings with
100 runs per setting.
%
After sampling events, we sample from the sensors
%at the corresponding location,
according to their false positive and false negative probabilities.  
\hide{For each of
the $i$ events which the sensor might detect, we consider the probability
$\mathit{fn}$ of a false negative, so the probability of the sensor firing is $1
- \mathit{fn}^i$. If none of the sensed events occurred, the sensor will fire
according to its false positive probability.
}

@@

\textbf{Configurations.}
\label{sec-5-1}
%% Explain how we generate settings, events, benign events, and sensor
%% reports.
% Generation of each experimental run of \mifd\ is governed by a number
% of experimental parameters.  
% Over any particular experiment, we will vary one or two parameters,
% and set others to values typical for an \ids.
In each of our experiments, we start with a configuration that reflects the
realities of IDS fusion, and whose parameter values are broadly consistent with
the simplifying assumptions of \zplus.
Configuration parameters may be either constants or parameterized random
variables.
We then vary some configuration parameter in ways which degrade
MIFD's performance. These experiments show us MIFD's performance under best-case conditions, and
how that performance degrades.
\begin{compactitem}
\item \numEventProtosDetected\ --- The total number of events (both
  benign and attacks) to be
  detected by each sensor, a measure of the sensor's specificity.
  Initially we use a random variable which returns 1 80\%
  of the time, and 2 otherwise.
\item \sensorToAttackRatio\ --- The number of sensors which should
  detect each attack event, initially 3.
\item \sensorOverlap\ --- The number of types of attack to be detected by each
  sensor, specified as a probability checked when deciding whether to
  associate an additional attack type with a sensor, initially 0.2.
\item \fpKappa\ --- The qualitative probability of
  false-positives for the sensors. We choose these to be 1
  in configurations where it takes 3 reports to rank a hypothesis likely, or 2
  where it takes only 2.
% 
%% Commenting out this bullet point --- not really relevant.
%
% \item $\kappa(\text{false-negative})$ --- MIFD does not currently reason about false negatives;
%  we take this probability to be whatever
%  corresponds to $\kappa = 2$ in the configuration.
%
\item \attackKappa\ and \benignKappa\ --- The qualitative
  probabilities of attack and benign events, initially
  2 and 1 respectively.

@@

\end{compactitem}
%
% The values of \sensorToAttackRatio\ and the various \emph{kappa}s are
% interrelated. A sensible scenario should respect the orderings
% \begin{align}
% \end{align}
% or in other words,
% \begin{compactitem}
% \item A single sensor false positive is less surprising than an
%   attack event.
% \item A benign event is also less surprising than an attack event.
% \item An attack event is less surprising than false positives from all
%   of the sensors which detect that event.
% \end{compactitem}
%
%% (run-config ...)
%%  > (make-configuration-settings configuration)
%%  > (make-setting :configuration configuration)
%%  > (make-instance 'setting :configuration configuration ...)
%%    :after (initialize-instance ((obj setting)) ...)
%%             > (make-attack-events num-attacks attack-kappa)
%%                 ;; just the attack protos
%%    (setf (slot-value setting 'sensors) (make-sensors setting))
%%                 ;; sensor and benign protos
%%    (stratus-simulator:init-random-mifd-world setting num-locations
%%    ...)
%
%
\textbf{Settings.}
When creating a setting, we first consult
the \numAttacks\ parameter to determine the number of attack prototypes.
Each prototype is associated with an
occurrence probability given by (or sampled for each prototype from)
\attackKappa. The number of sensors is not fixed, but instead depends
on several configuration parameters. For each attack, we keep track of
the number of sensors which we must assign to detect it; this value is
initially set from \sensorToAttackRatio.  We create new sensors as
long as any attack requires assignment to an additional sensor. A new
sensor is initially assigned a total number of events which it will
detect from \numEventProtosDetected, and some set of attack
events. Each sensor is assigned at least one attack, plus
additional attacks up to its total number of events as consecutive
samples of a Bernoulli random variable with success probability
\sensorOverlap\ return 1.  After all sensor prototypes are
created and assigned attacks, we assign benign events to the sensors
to reach their total number of events.
%% \begin{itemize}
%% \item If we are given \numBenignEvents, then we create that total
%%   number of benign event prototypes, and distribute them among the
%%   sensors using a \sensorToBenignRatio\ which allows each sensor's
%%   total event count to be satisfied.
%% \item
%%   If we are given a \sensorToBenignRatio, then we create the
%%   necessary number of benign event prototypes to satisfy both
%%   \sensorToBenignRatio\ and each sensor's total event count.
%% \end{itemize}
We create enough benign event prototypes to satisfy
both \sensorToBenignRatio\ and each sensor's total event count, and
distribute them randomly among the sensors according to their total
event counts. 
% Finally we generate a network with the number of nodes
% determined by \numLocations.

%% (nondegenerate-simulate-step setting ...)
%%   > (simulate-step setting)
%%   > (events-for-location setting location)
%%     (reports-for-events setting location events)
% From the setting we can generate the ground truth aspect for an actual
% \mifd\ run: actual attack and benign events, plus the sensor reports
% which those events (or false positives) trigger. We proceed by
% considering each location of the network in turn.  We create the
% events which occur at each location by considering each attack and
% benign prototype. For each prototype, if a sample of a Bernoulli
% random variable whose success probability is the occurrence
% probability of the prototype is 1, then the ground truth aspect will
% contain a sensor report with that prototype and location.
% %
% For sensor reports at each location, we consider each sensor report
% prototype of the setting. If a prototype indicates that the sensor can
% detect one of the events which we have generated for that location,
% then we may have a report based on the events; otherwise we may have
% report because of a false positive.
% \begin{compactitem}
% \item If we have detectable events, then for each one we sample a
%   Bernoulli random variable whose success probability is the false
%   negative probability of the report prototype. If \emph{any} of these
%   samples are 0 --- that is, if for any event we do \emph{not} have a
%   false negative --- then the ground truth aspect will contain a
%   sensor report with that prototype and location.
% \item If there are no events at a location which could trigger a
%   report of that prototype, then we consult a Bernoulli random
%   variable whose success probability is the false positive probability
%   of the sensor report prototype; if it returns 1, then the ground
%   truth aspect will contain a sensor report with that prototype and
%   location.
% \end{compactitem}

>>>>

% DONE: Can you give the reader a sense of the distinction between
% configuration, setting, and run? The notion of "run" is relatively clear, but
% not the other two.  Something like "For example, a configuration might specify
% how a set of attacks is generated, and the set of sensors. A particular
% setting would involve a particular set of possible attacks, a particular set
% of sensors, and a particular set of sensing relations between attacks and
% sensors."
%
% FIXME - Robert, in the marked-up PDF you suggested discussing
% locations, but earlier we'd decided to prune that discussion, since
% we didn't end up using different locations.
% Right, but here how will the reader know that we are going to sample from the
% same event multiple times?  The fact is that the whole notion of "run" as
% distinct from "location" is broken, so we must make the best of it that we
% can.  Please have a whack at this.
For each experiment we create a \emph{configuration},
a set of parameters specifying how a set of events and sensors, a
\emph{setting}, is generated.  A
setting comprises a set of possible attacks, a set
of sensors, and a set of sensing relations between attacks and
sensors.
In each \emph{run} of a setting we sample %, for each location,
% some number of 
from the attack and benign events.
% These are
% Bernoulli random variables; we discuss setting their probabilities
% below.
In each experiment, for each configuration, we generated 1,000 settings
and conducted 100 runs per setting.
%
After sampling events, we sample from the sensors
%at the corresponding location,
according to their false positive and false negative probabilities.  
\hide{For each of
the $i$ events which the sensor might detect, we consider the probability
$\mathit{fn}$ of a false negative, so the probability of the sensor firing is $1
- \mathit{fn}^i$. If none of the sensed events occurred, the sensor will fire
according to its false positive probability.
}

@@

\textbf{Configurations.}
\label{sec-5-1}
%% Explain how we generate settings, events, benign events, and sensor
%% reports.
% Generation of each experimental run of \mifd\ is governed by a number
% of experimental parameters.  
% Over any particular experiment, we will vary one or two parameters,
% and set others to values typical for an \ids.
In each of our experiments, we start with a configuration that reflects the
realities of IDS fusion, and whose parameter values are broadly consistent with
the simplifying assumptions of \zplus.
Configuration parameters may be either constants or parameterized random
variables.
We then vary some configuration parameter in ways which degrade
MIFD's performance. These experiments show us MIFD's performance under best-case conditions, and
how that performance degrades.
\begin{compactitem}
\item \numEventProtosDetected\ --- The total number of event types (both
  benign and attacks) to be
  detected by each sensor, a measure of the sensor's specificity. 
  Initially we use a random variable which returns 1 80\%
  of the time, and 2 otherwise.
\item \sensorToAttackRatio\ --- The number of sensors which should
  detect each attack event, initially 3.
\item \sensorOverlap\ --- The number of types of attack to be detected by each
  sensor, specified as a probability checked when deciding whether to
  associate an additional attack type with a sensor, initially 0.2.
\item \fpKappa\ --- The qualitative probability of
  false-positives for the sensors. We choose these to be 1
  in configurations where it takes 3 reports to rank a hypothesis likely, or 2
  where it takes only 2.
% 
%% Commenting out this bullet point --- not really relevant.
%
% \item $\kappa(\text{false-negative})$ --- MIFD does not currently reason about false negatives;
%  we take this probability to be whatever
%  corresponds to $\kappa = 2$ in the configuration.
%
\item \attackKappa\ and \benignKappa\ --- The qualitative
  probabilities of attack and benign events, initially
  2 and 1 respectively.

@@

\end{compactitem}
%
% The values of \sensorToAttackRatio\ and the various \emph{kappa}s are
% interrelated. A sensible scenario should respect the orderings
% \begin{align}
% \end{align}
% or in other words,
% \begin{compactitem}
% \item A single sensor false positive is less surprising than an
%   attack event.
% \item A benign event is also less surprising than an attack event.
% \item An attack event is less surprising than false positives from all
%   of the sensors which detect that event.
% \end{compactitem}
%
%% (run-config ...)
%%  > (make-configuration-settings configuration)
%%  > (make-setting :configuration configuration)
%%  > (make-instance 'setting :configuration configuration ...)
%%    :after (initialize-instance ((obj setting)) ...)
%%             > (make-attack-events num-attacks attack-kappa)
%%                 ;; just the attack protos
%%    (setf (slot-value setting 'sensors) (make-sensors setting))
%%                 ;; sensor and benign protos
%%    (stratus-simulator:init-random-mifd-world setting num-locations
%%    ...)
%
%
\textbf{Settings.}
When creating a setting, we first consult
the \numAttacks\ parameter to determine the number of attack types.
Each type is associated with an
occurrence probability dictated by the 
\attackKappa parameter. The number of sensors is not fixed, but instead depends
on several configuration parameters. For each attack, we keep track of
the number of sensors which we must assign to detect it; this value is
initially set from \sensorToAttackRatio.  We create new sensors as
long as any attack requires assignment to an additional sensor. A new
sensor is initially assigned a total number of events which it will
detect from \numEventProtosDetected, and some set of attack
events. Each sensor is assigned at least one attack, plus
additional attacks up to its total number of events based on a 
Bernoulli random variable with parameter
\sensorOverlap.  After all sensor types are
created and assigned attacks, we assign benign events to the sensors
to reach their total number of events.
%% \begin{itemize}
%% \item If we are given \numBenignEvents, then we create that total
%%   number of benign event types, and distribute them among the
%%   sensors using a \sensorToBenignRatio\ which allows each sensor's
%%   total event count to be satisfied.
%% \item
%%   If we are given a \sensorToBenignRatio, then we create the
%%   necessary number of benign event types to satisfy both
%%   \sensorToBenignRatio\ and each sensor's total event count.
%% \end{itemize}
We create enough benign event types to satisfy
both \sensorToBenignRatio\ and each sensor's total event count, and
distribute them randomly among the sensors according to their total
event counts. 
% Finally we generate a network with the number of nodes
% determined by \numLocations.

%% (nondegenerate-simulate-step setting ...)
%%   > (simulate-step setting)
%%   > (events-for-location setting location)
%%     (reports-for-events setting location events)
% From the setting we can generate the ground truth aspect for an actual
% \mifd\ run: actual attack and benign events, plus the sensor reports
% which those events (or false positives) trigger. We proceed by
% considering each location of the network in turn.  We create the
% events which occur at each location by considering each attack and
% benign type. For each type, if a sample of a Bernoulli
% random variable whose success probability is the occurrence
% probability of the type is 1, then the ground truth aspect will
% contain a sensor report with that type and location.
% %
% For sensor reports at each location, we consider each sensor report
% type of the setting. If a type indicates that the sensor can
% detect one of the events which we have generated for that location,
% then we may have a report based on the events; otherwise we may have
% report because of a false positive.
% \begin{compactitem}
% \item If we have detectable events, then for each one we sample a
%   Bernoulli random variable whose success probability is the false
%   negative probability of the report type. If \emph{any} of these
%   samples are 0 --- that is, if for any event we do \emph{not} have a
%   false negative --- then the ground truth aspect will contain a
%   sensor report with that type and location.
% \item If there are no events at a location which could trigger a
%   report of that type, then we consult a Bernoulli random
%   variable whose success probability is the false positive probability
%   of the sensor report type; if it returns 1, then the ground
%   truth aspect will contain a sensor report with that type and
%   location.
% \end{compactitem}

### 2015-03-12 intro.tex
Minor edits to the text.
One note: as I read this, I realized that the reader doesn't need to
know that we have event and sensor PROTOTYPES instead of TYPES. That's
an implementation detail that comes from an oddity of PRAXIS. So I
suppressed mention of "prototypes."

We experimentally analyze the accuracy of the \zplus qualitative
probability scheme when used for diagnosis and information fusion.
In cyber defense, an ``intrusion detection system (IDS) is a device or
software application that monitors network or system activities for malicious
activities or policy violations''~\cite{wiki:ids}.
In previous work we developed a technique for IDS
fusion, deployed in the Scyllarus system~\anoncite{goldman:09Scyllarus} and its
successor MIFD~\anoncite{stratus:saso12}.
These systems fuse together reports from multiple, heterogeneous IDSes,
hypothesizing underlying events to explain those reports, and assessing the events'
likelihood, to detect cyber attacks.
Their likelihood assessment is based on \zplus, a qualitative
abstraction of probability theory~\cite{Goldszmidt:96}.
Scyllarus has been extensively tested in real networks, using both
real and synthetic data, and has shown its ability to accurately fuse reports
from extremely noisy sensors. 
We also
drastically reduce false alarm rates, the bane of intrusion
detection systems, reducing the flood of incoming reports by multiple
orders of magnitude.
Unfortunately, it is difficult to draw crisp, \emph{general} conclusions based on
field evaluations alone.
We need to demonstrate that the results are due to features of the algorithm,
not simply artifacts of details of the test network, traffic, and
attacks.  This problem is particularly acute in the area of intrusion detection,
as we explain below.
% This is a general problem of evaluating IDSes, which we discuss
% further below.
%

@@

To the best of our knowledge, ours is the only empirical work to explore the
\emph{accuracy} of reasoning with \zplus.
Of about 250 works citing Goldszmidt and Pearl's 1996
work~\cite{google-scholar-query}, none of the few which detail
applications of \zplus\ reasoning conduct such investigations.
%
Minock and Kraus~\shortcite{minock-kraus-LIAI02} investigate the
\emph{efficiency} of an implementation of \zplus,
\hide{restricted to specific classes of
Horn theories,}
but not its accuracy.
% Minock and Kraus compile queries over Horn and q-Horn clauses to an
% implementation of \zplus, but they report only ``initial performance
% results'' for partition construction to illustrate the tractability of
% their approach~\shortcite{minock-kraus-LIAI02}.
%
% Fujimoto and Matsuzawa use qualitative rankings similar to \zplus\ to
% categorize subjective evaluations in
% text~\shortcite{fujimoto-matsuzawa-newgencomp99,fujimoto-WIIAT2010};
% their system predicts reactions to advertising messages.  Fujimoto's
% experiment~\shortcite{fujimoto-WIIAT2010} is directed towards a
% regression analysis, examining prediction accuracy on a set of survey
% responses.
%
% We do not question the correctness of those experiments, but
% the substitution of qualitative probabilities for actual values should
% not be made without confirming the appropriateness of the
% approximation for the particular domain.  We believe that the
% experiments we report here are the first such rigorous investigation
% of the suitability of \zplus\ for a specific application.

>>>>

We experimentally analyze the accuracy of the \zplus qualitative
probability scheme when used for diagnosis and information fusion.
In cyber defense, an ``\ldots intrusion detection system (IDS) is a device or
software application that monitors network or system activities for malicious
activities or policy violations''~\cite{wiki:ids}.
In previous work we developed a technique for IDS
fusion, deployed in the Scyllarus system~\anoncite{goldman:09Scyllarus} and its
successor MIFD~\anoncite{stratus:saso12}.
These systems fuse together reports from multiple, heterogeneous IDSes,
hypothesizing underlying events to explain those reports, and assessing the events'
likelihood, to detect cyber attacks.
Their likelihood assessment is based on \zplus, a qualitative
abstraction of probability theory~\cite{Goldszmidt:96}.
Scyllarus has been extensively tested in real networks, using both
real and synthetic data, and has shown its ability to accurately fuse reports
from extremely noisy sensors. 
We also
dramatically reduce false alarm rates, the bane of intrusion
detection systems, reducing the flood of incoming reports by multiple
orders of magnitude.
Unfortunately, we cannot draw crisp, \emph{general} conclusions based on
field evaluations alone.
We need to demonstrate that the results are due to features of the algorithm,
not simply artifacts of details of the test network, traffic, and
attacks.  This problem is particularly acute in the area of intrusion detection,
as we explain below.
% This is a general problem of evaluating IDSes, which we discuss
% further below.
%

@@

To the best of our knowledge, ours is the only empirical work to explore the
\emph{accuracy} of reasoning with \zplus.
Of about 250 works citing Goldszmidt and Pearl's 1996
work~\cite{google-scholar-query}, none of the few which detail
applications of \zplus\ reasoning (as opposed to theoretical investigations of
the logic) conduct such investigations.
%
Minock and Kraus~\shortcite{minock-kraus-LIAI02} investigate the
\emph{efficiency} of an implementation of \zplus,
\hide{restricted to specific classes of
Horn theories,}
but not its accuracy.
% Minock and Kraus compile queries over Horn and q-Horn clauses to an
% implementation of \zplus, but they report only ``initial performance
% results'' for partition construction to illustrate the tractability of
% their approach~\shortcite{minock-kraus-LIAI02}.
%
% Fujimoto and Matsuzawa use qualitative rankings similar to \zplus\ to
% categorize subjective evaluations in
% text~\shortcite{fujimoto-matsuzawa-newgencomp99,fujimoto-WIIAT2010};
% their system predicts reactions to advertising messages.  Fujimoto's
% experiment~\shortcite{fujimoto-WIIAT2010} is directed towards a
% regression analysis, examining prediction accuracy on a set of survey
% responses.
%
% We do not question the correctness of those experiments, but
% the substitution of qualitative probabilities for actual values should
% not be made without confirming the appropriateness of the
% approximation for the particular domain.  We believe that the
% experiments we report here are the first such rigorous investigation
% of the suitability of \zplus\ for a specific application.

### 2015-03-12 scyllarus-and-mifd.tex
Minor edits to the text.
One note: as I read this, I realized that the reader doesn't need to
know that we have event and sensor PROTOTYPES instead of TYPES. That's
an implementation detail that comes from an oddity of PRAXIS. So I
suppressed mention of "prototypes."

Our original system Scyllarus
%% TODO Restore for QR'15
% was an open system which 
performed IDS
fusion on a diverse set of third party \idses~\anoncite{goldman:discex2001,goldman:discex2001}.
Its successor
MIFD (Model-based Intrusion Fusion and Detection),
enhances Scyllarus
%% TODO Restore for QR'15
% as part of the STRATUS system's integration of
 to integrate
IDS fusion into a comprehensive suite for cyber defense
\anoncite{thayer13:_compar_strat_tactic_respon_cyber_threat}.
For simplicity we refer to our \ids fusion system as ``MIFD,'' but
except where we specify otherwise, our account applies to Scyllarus as well.

The first tasks performed by MIFD are
\emph{collection} and \emph{translation}: IDS reports
must be collected and presented to MIFD, and
must be translated into a common
sensor report data structure for processing.
We will not discuss these less interesting
preliminaries, except to mention that they
involve a substantial amount of data rectification
%% TODO Restore for QR'15
 to address the lack of standardization for data formats and event taxonomies.
%.  There is no widely-accepted
%common data format (the proposed IDMEF standard~\cite{IDMEF} has not caught on), nor
%is there an accepted taxonomy for events or event roles (e.g., the widely used ``source''
%and ``destination'' are not consistently applied).
%These problems are exacerbated by the fact that intrusion detection ``systems''
%are often simply a casually-assembled anthology of rules by different authors running in the same
%framework.

Core fusion inference begins with
\textbf{Clustering:} From the
sensor reports, MIFD generates \emph{event hypotheses} to represent the
underlying events of interest that may have caused the sensor reports.  A
sensor report that provides evidence for an event hypothesis is a
\emph{supporter}.
The clustering process constructs a
directed graph of event hypothesis and sensor report nodes forming a
Bayesian belief network.

@@

\ids rules detect indirect features of intrusions, but they typically
report only the intrusion that their designer believes has caused the feature.
This creates a number of problems, first and foremost that the
sensors have high false positive rates.
For example, a local software update server
may look like an attacker performing reconnaissance.
Worse, some false positives are correlated: the normal
activity of a print server may be mis-detected as a scan by several
different sensors.
To handle simple sensor failures, MIFD associates with each sensor a false
positive likelihood.
But to address the correlated false positives such as the scanning example, we
introduce benign events, that can compete with attack and (simple) false
positive explanations for results.\footnote{And where possible, we try to introduce sensors that
can specifically detect the benign events. 
%% TODO Restore for QR'15
%To return to our earlier example, we
%might introduce a ``print server sensor'' to help disambiguate detections. We do
%not address this issue in our experiments.
}
% FIXME: Add a figure 2(c) -- or separately numbered figure -- where we have a
% benign event, explaining two of the sensor reports, as well as the two attack
% events. 
Another problem with indirect sensing is that the sensors are often imprecise:
the features that they detect often occur not only in the absence of the
advertised intrusion, but in the presence of \emph{other} intrusions with
similar characteristics.

@@

The final inference step is \textbf{Assessment:} MIFD assigns a degree of belief
to each of the event hypotheses.
MIFD does this by performing Bayesian updating on the belief network constructed
by the clustering process. 
%% As we discuss below, MIFD does this using the \zplus qualitative probability
%% scheme rather than standard probability theory.
%%
Scyllarus and MIFD are intended as long-running online services, not batch processes.
Scyllarus has run for months at a time.
One major requirement for such long-term operation is that the
inference steps must be performed cyclically.  MIFD repeatedly reads
reports and clusters them, periodically interrupting that
cycle to perform assessment.
%% FIXED Although this next sentence is a true fact, I'm not sure it
%% belongs here --- it makes assessment efficient, but that's not per
%% se an inherent requirement for long-term operation.
% Moreover, in general the Bayesian belief network
% is not fully connected, so for efficiency MIFD assesses each connected
% component individually.
% \footnote{%
%   An additional essential requirement for long-term operation, not
%   relevant to the remainder of this presentation, is some form of
%   garbage collection: old event hypotheses and sensor reports
%   guaranteed no longer to be relevant to subsequent report must be
%   removed from working memory.}

>>>>

Our original system Scyllarus
%% TODO Restore for QR'15
% was an open system which 
performed IDS
fusion on a diverse set of third party \idses~\anoncite{goldman:discex2001,goldman:discex2001}.
Its successor,
MIFD (Model-based Intrusion Fusion and Detection),
enhances Scyllarus
%% TODO Restore for QR'15
% as part of the STRATUS system's integration of
 to integrate
IDS fusion into a comprehensive suite for cyber defense
\anoncite{thayer13:_compar_strat_tactic_respon_cyber_threat}.
For simplicity we refer to our \ids fusion system as ``MIFD,'' but
except where we specify otherwise, our account applies to Scyllarus as well.

The first tasks performed by MIFD are
\emph{collection} and \emph{translation}: IDS reports
must be collected and presented to MIFD, and
must be translated into a common
sensor report data structure for processing.
We will not discuss these less interesting
preliminaries, except to mention that they
involve a substantial amount of data rectification
%% TODO Restore for QR'15
 to address the lack of standardization in data formats and event taxonomies.
%.  There is no widely-accepted
%common data format (the proposed IDMEF standard~\cite{IDMEF} has not caught on), nor
%is there an accepted taxonomy for events or event roles (e.g., the widely used ``source''
%and ``destination'' are not consistently applied).
%These problems are exacerbated by the fact that intrusion detection ``systems''
%are often simply a casually-assembled anthology of rules by different authors running in the same
%framework.

{\ids} fusion proper begins with
\textbf{Clustering.} From the
sensor reports, MIFD generates \emph{event hypotheses} to represent the
underlying events of interest that may have caused the sensor reports.  A
sensor report that provides evidence for an event hypothesis is a
\emph{supporter}.
The clustering process constructs a
directed graph of event hypothesis and sensor report nodes forming a
Bayesian belief network.

@@

\ids rules detect indirect features of intrusions, but they typically
report only the intrusion that their designer believes has caused the feature.
This creates a number of problems, first and foremost that the
sensors have high false positive rates.
For example, a local software update server
may look like an attacker performing reconnaissance.
Worse, some false positives are correlated: the normal
activity of a print server may be mis-detected as a scan by several
different sensors.
To handle simple sensor failures, MIFD associates with each sensor a false
positive likelihood.
But to address the correlated false positives such as the scanning example, we
introduce benign events, that can compete with attack and (simple) false
positive explanations for results.\footnote{Where possible, we try to introduce sensors that
can specifically detect the benign events. 
%% TODO Restore for QR'15
%To return to our earlier example, we
%might introduce a ``print server sensor'' to help disambiguate detections. We do
%not address this issue in our experiments.
}
% FIXME: Add a figure 2(c) -- or separately numbered figure -- where we have a
% benign event, explaining two of the sensor reports, as well as the two attack
% events. 
Another problem with indirect sensing is that the sensors are often imprecise:
the features that they detect often occur not only in the absence of the
advertised intrusion, but in the presence of \emph{other} intrusions with
similar characteristics.

@@

The second inference step is \textbf{Assessment,} in which MIFD assigns a degree of belief
to each of the event hypotheses.
MIFD does this by performing Bayesian updating on the belief network constructed
by the clustering process. 
%% As we discuss below, MIFD does this using the \zplus qualitative probability
%% scheme rather than standard probability theory.
%%

Note that Scyllarus and MIFD are intended as long-running online services, not batch processes.
Scyllarus has run for months at a time.
One major requirement for such long-term operation is that the
inference steps must be performed cyclically.  MIFD repeatedly reads
reports and clusters them, periodically interrupting that
cycle to perform assessment.
Scyllarus also periodically flushes older events and reports out of working
memory (they persist in a database).

%% FIXED Although this next sentence is a true fact, I'm not sure it
%% belongs here --- it makes assessment efficient, but that's not per
%% se an inherent requirement for long-term operation.
% Moreover, in general the Bayesian belief network
% is not fully connected, so for efficiency MIFD assesses each connected
% component individually.
% \footnote{%
%   An additional essential requirement for long-term operation, not
%   relevant to the remainder of this presentation, is some form of
%   garbage collection: old event hypotheses and sensor reports
%   guaranteed no longer to be relevant to subsequent report must be
%   removed from working memory.}

### 2015-03-12 zplus.tex
No longer need to divide anon. from other citations.
Minor edits to the text.
One note: as I read this, I realized that the reader doesn't need to
know that we have event and sensor PROTOTYPES instead of TYPES. That's
an implementation detail that comes from an oddity of PRAXIS. So I
suppressed mention of "prototypes."

MIFD
follows Pearl's exhortation~\shortcite{Pearlbook} that the core of probability theory is
the patterns of inference it enables (explaining away, combining forward causal
inference and evidential reasoning, etc.), rather than precise numerical
calculations.  He recommends these patterns be used even when precise numbers are
not available, which MIFD does.
%% TODO Restore for QR'15
%As outlined above,
MIFD treats
\ids fusion as an abductive problem, formalized using Bayes nets.
But precise values for the parameters of these Bayes nets
%% TODO Restore for QR'15
% --- probabilities of attacks, false positives, false negatives, etc. ---
are \emph{never} available to us: populations of
attacks and attackers change, different networks have little in common with each other, distributions are
non-stationary, etc.  Moreover these difficulties are not solvable via machine learning:
Sommer and Paxson~\shortcite{Sommer10outsidethe} discuss the challenges \ids pose to ML.

@@

There are a number of efficient algorithms for finding the posterior distributions of Bayesian networks, 
conditional on observations of some of the random variables. These algorithms may readily be adapted to 
provide posterior $\kappa$ rankings instead of probabilities.
MIFD translates Bayes networks into an
%% DONE Charniak:88 is Charniak and Goldman --- withhold this one too? No --
%% it's not obviously tied to this paper, so we don't need to worry about it
%% breaching anonymity. [2014/08/24:rpg]
ATMS~\cite{Forbus+DeKleer:1993}; see \cite{Charniak:88,Poole:93,provan-ijcai89}
and \anoncite{goldman:09Scyllarus} for this encoding.
This implementation is not especially efficient, but we have found
it is I/O which dominates runtime, and we apply several optimizations to
handle common special cases.
%% TODO Restore for QR'15
%To date (somewhat to our disappointment!) optimizing the general \zplus
%inference method has not been key to system performance.
%
% To date, finding the optimal inference method
% % in terms of runtime
% has been
% less important than finding an algorithm easy to modify for \zplus.

>>>>

MIFD
follows Pearl's exhortation~\shortcite{Pearlbook} that the core of probability theory is
the patterns of inference it enables (explaining away, combining forward causal
inference and evidential reasoning, etc.), rather than precise numerical
calculations.  He recommends these patterns be used even when precise numbers are
not available.
%% TODO Restore for QR'15
%As outlined above,
MIFD treats
\ids fusion as an abductive problem, formalized using Bayes nets.
But precise values for the parameters of these Bayes nets
%% TODO Restore for QR'15
% --- probabilities of attacks, false positives, false negatives, etc. ---
are \emph{never} available to us: populations of
attacks and attackers change, different networks have little in common with each other, distributions are
non-stationary, etc.  Moreover these difficulties are not solvable via machine learning:
Sommer and Paxson~\shortcite{Sommer10outsidethe} discuss the challenges \idses pose to ML.

@@

There are a number of efficient algorithms for finding the posterior distributions of Bayesian networks, 
conditional on observations of some of the random variables. These algorithms may readily be adapted to 
provide posterior $\kappa$ rankings instead of probabilities.
MIFD translates Bayes networks into an
%% DONE Charniak:88 is Charniak and Goldman --- withhold this one too? No --
%% it's not obviously tied to this paper, so we don't need to worry about it
%% breaching anonymity. [2014/08/24:rpg]
ATMS~\cite{Forbus+DeKleer:1993}; see
%% For anonymous:
%\cite{Charniak:88,Poole:93,provan-ijcai89}
%and \anoncite{goldman:09Scyllarus}
%% De-anonymized:
\cite{Charniak:88,Poole:93,provan-ijcai89,goldman:09Scyllarus}
%
for this encoding.
This implementation is not especially efficient, but we have found
it is I/O which dominates runtime, and we apply several optimizations to
handle common special cases.
%% TODO Restore for QR'15
%To date (somewhat to our disappointment!) optimizing the general \zplus
%inference method has not been key to system performance.
%
% To date, finding the optimal inference method
% % in terms of runtime
% has been
% less important than finding an algorithm easy to modify for \zplus.
